<!-- %:::% .common h1 begin %:::% -->
# Associations between the duration and quality of sleep of pregnant women in the third trimester with the duration of labor
<!-- %:::% .common h1 end %:::% -->

```{r}
#| label: setup
#| include: false

source(here::here("R/quarto-setup.R"))
```

```{r}
#| echo: false
#| output: asis

rutils:::quarto_status(
  type = "drafting",
  of_what = "of this thesis",
  latex_parskip = "\\microskip"
  )
```

::: {.callout-note}
## Target journal

1. [Chronobiology International](https://www.tandfonline.com/action/authorSubmission?show=instructions&journalCode=icbi20) ([IF 2022: 2.8/JCR](https://jcr.clarivate.com/jcr/) | [A1/2017-2020](https://sucupira.capes.gov.br/sucupira/public/consultas/coleta/veiculoPublicacaoQualis/listaConsultaGeralPeriodicos.jsf)).
2. [Journal of Biological Rhythms](https://journals.sagepub.com/author-instructions/JBR) ([IF 2022: 3.5/JCR](https://jcr.clarivate.com/jcr/) | [A2/2017-2020](https://sucupira.capes.gov.br/sucupira/public/consultas/coleta/veiculoPublicacaoQualis/listaConsultaGeralPeriodicos.jsf)).
:::

::: {.callout-note}
The following study was performed by Alícia Rafaelly Vilefort Sales (__ARVS__), Daniel Vartanian (__DV__), Maria Augusta Medeiros de Andrade (__MAMA__), Mario Pedrazzoli (__MP__) and Christiane Borges do Nascimento Chofakian (__CBNC__).

\microskip

__ARVS__, __DV__ and __MAMA__ contributed to the study's  design. __ARVS__ implemented the study, performed the statistical analysis, and authored the manuscript. __CBNC__ and __MP__ served as scientific advisors. All authors participated in discussions about the results and contributed to the final manuscript revision.

\microskip

_Future reference_: Sales, A. R. V., Vartanian, D., Andrade, M. A. M., Pedrazzoli, M., & Chofakian, C. B. N. (2024). Associations between the duration and quality of sleep of pregnant women in the third trimester with the duration of labor. _Chronobiology International_.
:::
